---
title: "Part 3: Data visualizations"
author: Rebecca Ross, Genoveva Gonzalez-Mirelis
output: 
  html_notebook: 
    toc: yes
---
## Introduction

In this Notebook we make an alternative version of Figure S1, which was done in GIS for the purpose of publication.

## Libraries
```{r, message=FALSE, warning = FALSE}
library(rgdal)
#library(dplyr)
library(sf)
library(sp)
library(raster)
library(rasterVis)
library(RColorBrewer)
library(ggplot2)
library(scales)
library(vegan)
setwd("/home/rstudio/fmars-2020-496688/")
```


## Import a Basemap

Bathymetric contours were derived from GEBCO bathymetry (GEBCO Compilation Group (2020) GEBCO 2020 Grid (doi:10.5285/a29c5465-b138-234d-e053-6c86abc040b9)). In the Data folder we provide this contour shapefile, already made.

The land layer was clipped from a world dataset downloaded from marineregions.org (Flanders Marine Institute (2018). Maritime Boundaries Geodatabase: Maritime Boundaries and Exclusive Economic Zones (200NM), version 10. Available online at http://www.marineregions.org/. https://doi.org/10.14284/312). In the Data folder we provide this shapefile already clipped.
```{r, message=FALSE, warning = FALSE,comment=NA}

baslay_path <- "Data/BaseLayers"

bathy.line <- readOGR(dsn = baslay_path, layer = "gebbathy_no_top_conts_utm33n")
land.mask<-readOGR(dsn=baslay_path, layer = "europe_africa_utm33n")
#sp::plot(bathy.line) #check

in.shp <- readOGR(dsn = baslay_path, layer = "response_spatial")

#hotspots <- readOGR(dsn = "Outputs/ModelPredictions", layer = "high_density_areas")

```

## Fortifications

Used to make shapefiles suitable for ggplotting.

```{r, message=FALSE, warning = FALSE}

my.shp<-fortify(as.data.frame(in.shp))
ax <-my.shp$coords.x1 # using these to set the zoom x and y limits of the plot
ay <-my.shp$coords.x2

my.bathy<-fortify(bathy.line)

my.land<-fortify(land.mask)

#my.hotspots <- fortify(hotspots)

```

## Plot the Basemap

```{r, message=FALSE, warning = FALSE}
# making the basemaps0
map.bathy<- ggplot()+
  coord_fixed(xlim=c(min(ax),max(ax)), ylim=c(min(ay), max(ay)))+
  geom_path(data=my.bathy, colour="grey", aes(x=long, y=lat, group=group))
 
map.bathy.z<- ggplot()+
  coord_fixed(xlim=c(492468.9345+15000, 788802.8605-15000), ylim=c(7757810.4574+15000,8047794.3707-15000))+
  geom_path(data=my.bathy, colour="grey", aes(x=long, y=lat, group=group))

map.bathy.z.o<- ggplot()+
  coord_fixed(xlim=c(492468.9345-1600000, 788802.8605+500000), ylim=c(7757810.4574-3000000,8047794.3707+500000))+
  geom_path(data=my.bathy, colour="white", aes(x=long, y=lat, group=group))

map.base<-map.bathy +
  geom_polygon(data=my.land, fill="#f0f0f0", colour="grey", aes(x=long, y=lat, group=group))

map.base.z<-map.bathy.z +
  geom_polygon(data=my.land, fill="#f0f0f0", colour="grey", aes(x=long, y=lat, group=group))

#map.base.z #check
  
```

## Plot the Shapefile map

```{r, message=FALSE, warning = FALSE}

# map with shapefile 

# make the PA column first

my.shp$Hbtt_dm <- decostand(my.shp$QAbu, "pa", na.rm =TRUE)

map.shp<-map.base + 
  geom_point(data=my.shp, aes(ax, ay, color=factor(Hbtt_dm)))+
  scale_color_manual(values=c("#d8b365","#5ab4ac"), 
                     name="Soft-bottom, deep-sea sponges",
                     labels=c("Absent", "Present"))+
  labs(x="Eastings", y="Northings")+
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.text.y = element_text(angle=90),
        legend.position="bottom",
        plot.caption = element_text(hjust = 0))

map.shp
 
```

##### Figure S2. Sampling stations colored according to the presence/absence of soft-bottom, deep-sea sponges

