---
title: "Part 4: Generation of model (spatial) predictions"
author: Genoveva Gonzalez-Mirelis
output: 
  html_notebook: 
    toc: yes
---

## Introduction

In this part we generate predictions from the selected models for the whole of the modelling area.

Inputs are: the models, and the predictor stack. The outputs are two rasters of model predictions.

## Libraries and paths

Load necessary libraries
```{r,comment=NA}

setwd("/home/rstudio/fmars-2020-496688/")
library(rgdal)

ifelse(!dir.exists(file.path("Outputs", "ModelPredictions")), dir.create(file.path("Outputs", "ModelPredictions")), FALSE)

```

## Load models
```{r,comment=NA}

load("Outputs/Models/probmodv2.RData")
model.prob <- proballall

load("Outputs/Models/densmodv2.RData")
model.dens <- densallall

```

## Predict

Declare factor variables. Write a function for later use.
```{r,comment=NA}

## Predict using the raster package function:

# list of factor levels

f1 <- levels(v$landscape)
f2 <- levels(v$terclass)

f <- list(f1,f2)

names(f) <-c("landscape","terclass")

predfun <- function(m, d, ...) predict(m, newdata=d, ...)

#pc <- raster::predict(pred, model, OOB=TRUE, factors=f, fun=predfun)

## same, in parallel

```

First, run the prediction function for probability model, in parallel.
```{r,comment=NA}

if("rgrass7" %in% (.packages())){
  detach("package:rgrass7", unload=TRUE)
}



raster::beginCluster()

pc.prob <- clusterR(pred, predict, args=list(model=model.prob, OOB=TRUE, factors=f, fun=predfun, index=1:2))

pc.prob <- pc.prob[[1]]

```

Then run for density model.
```{r,comment=NA}

beginCluster()

pc.dens <- clusterR(pred, predict, args=list(model=model.dens, OOB=TRUE, factors=f, fun=predfun, index=1:2))

pc.dens <- pc.dens[[1]]



```

## Plot the predictions

Probability
```{r}
plot(pc.prob)
```

Density
```{r, message=FALSE}
plot(pc.dens)
```


## Outputs

Save the results.
```{r}
outdatapath <- "Outputs/ModelPredictions"

response.dens <- "Dens"
response.prob <- "Prob"

writeRaster(pc.prob,file.path(outdatapath, paste0("Prediction", response.prob, ".tif", collapse = NULL)), options="INTERLEAVE=BAND", overwrite=TRUE)


writeRaster(pc.dens,file.path(outdatapath, paste0("Prediction", response.dens, ".tif", collapse = NULL)), options="INTERLEAVE=BAND", overwrite=TRUE)

```


