---
title: "Part 5: Correlation between model predictions and generation of zone files"
author: Genoveva Gonzalez-Mirelis
output: 
  html_notebook: 
    toc: yes
---

## Introduction

In this part we calculate some measures of correlation between predictions from either model, to determine the level of agreement between the two models.

Th inputs include the two rasters of predictions generated in Part 4. The outputs include a raster of model correlations, as well as other values, which are described in the text of the accompanying paper.

Here we make an alternative (and more detailed) version of Figure 3.

## Correlation between models

### Libraries and data
```{r}
library(raster)
library(spatialEco)
library(rgdal)
library(smoothr)

setwd("/home/rstudio/fmars-2020-496688/")

pred_dens <- raster("Outputs/ModelPredictions/PredictionDens.tif")
pred_prob <- raster("Outputs/ModelPredictions/PredictionProb.tif")

```

### Calculate overall, and local Pearson correlation

Overall correlation
```{r}
overall_rho <- cor(values(pred_dens),values(pred_prob), use = "na.or.complete", method = "pearson")
overall_rho
```

Local correlation
```{r}
r.cor <- rasterCorrelation(pred_dens, pred_prob, s = 51, type = "pearson")
plot(r.cor)
```
##### Figure s3. Local correlation between prediction of density and prediction of probability of presence of soft-bottom, deep-sea sponges.

### Percent areas in bins

Plot the histogram of all values
```{r}
a <- values(r.cor)[which(!(is.nan(values(r.cor))|(is.na(values(r.cor)))))]

hist(a)

```

Percent area with correlation >0.2
```{r}
b <- a[which(a>=0.2)]
length(b)/length(a)
```

Percent area with correlation between 0 and 0.7
```{r}
b <- a[which(a<0.2 & a>-0.2)]
length(b)/length(a)
```

Percent area with negative correlation
```{r}
b <- a[which(a<(-0.2))]
length(b)/length(a)
```

Percent area with correlation >0.7
```{r}
b <- a[which(a>=0.7)]
length(b)/length(a)
```


### Save result

```{r}
outdatapath <- "Outputs/ModelPredictions"
writeRaster(r.cor,file.path(outdatapath, "pearson_correlation_51.tif"), options="INTERLEAVE=BAND", overwrite=TRUE)
```



